Reproducible and programmatic generation of neuroimaging visualisations

### 2022-03-31 manuscript.Rmd
figure compiled

---
title: "Reproducible and programmatic generation of neuroimaging visualisations"
author: "Sidhant Chopra & [coauthors]"
date: "09/03/2022"
output:
  word_document: default
  pdf_document: default
  html_document: default
editor_options:
  markdown:
    wrap: 72
---

@@

The `oro.nifti` package allows for the reading, writing, manipulation
and visualisation of voxel-level imaging data of either *nifti*,
*analyze* or *afni* formats. In particular, the `image`, `slice` and
`orthographic` functions allow users to display the desired number of
slices in the desired orientations, while also providing precise control
over overlaied images, colour scales, legend placement and other
aesthetic settings (Fig 1A). Smoothing and re-sampling of images is
sometimes required for visualisation purposes, which can be achieved via
the *`ANTsR`* /`extrantsr` or `fslr` packages.

@@

Statistical parameters can be mapped onto discreet cortical and
subcortical regions using the `ggseg` and `ggseg3d` package. These
packages flexibly generates aesthetic renderings of cortical,
subcortical and white matter ROIs in 2D and 3D (Fig 1D). Usually,
ROI-level interpretations are dependent on a standerdised atlas or
parcellation schemes the brain, accordingly this package and its add-ons
(`ggExtra`) provides a large array of commonly used atlases and some
functionality for users to contribute other atlases. Additionally, being
part of the 'Grammar of Graphics' frame work allows for integration with
packages such as `gganimate`, enabling users to animate and visualise
dynamic changes in spatial and temporal statistics across ROIs. The
`ciftiTools` package also allows for surface and voxel-level
visualisation of ROIs for the cortex and subcortex respectively (Fig
1E). Flexibly allowing CIFTI format atlases and ROIs to be displayed.

@@

-   The below figures will be combined into a multi-panel figure with
    the corresponding code underneath.

```{r}
#Figure 1A
#voxel 1

@@

library(neurobase)

@@

#Plot the effect on the brain template
ortho2(x = template, 
       y = effect, 
       crosshairs = F, 
       bg = "white", 
       NA.x = T, 
       col.y  = hotmetal(), 
       xyz = c(70,50,80),
       useRaster = T,
       ybreaks = seq(0.3, 1, length.out = 65), 
       ycolorbar = TRUE, 
       mfrow = c(1,3))

@@

```{r}
#Figure 1B
#voxel 2

library(neurobase)

library(scales)

@@

overlay(x=robust_window(dti), y=atlas, 
               plot.type = "single", 
               z=c(seq(25,110,5)), 
               col.y =  alpha(rainbow(300), 0.3),
               plane = "axial",  
               useRaster = T, 
               bg = "white",NA.x=T, 
               zlim.y =c(1,300))

@@

```{r, fig.cap="Snapshot", rgl=TRUE, format="png", eval=TRUE}
#vertex 1
#Figure 1B

library(rgl)

# Load the package and point to the Connectome Workbench
suppressPackageStartupMessages(library(ciftiTools))

@@

view_xifti_surface(cii, 
                   hemisphere = "left", 
                   view = "lateral", 
                   idx = 1,
                   colors = viridis::viridis(n = 100),
                   zlim = c(1,1.5), legend_embed = T)

@@

```{r, fig.cap="Snapshot", rgl=TRUE, format="png", eval=TRUE}
#Vertex 2
#Figure 1X - Remove white space

library("fsbrain")

subjects_dir = fsbrain::get_optional_data_filepath("subjects_dir")

@@

vis.subject.morph.native(subjects_dir = subjects_dir, 
                         subject_id = 'subject1', 
                         measure = 'thickness', 
                         hemi='both', 
                         views="t4",
                         draw_colorbar = "horizontal",
                         cortex_only = T, surface = "white",
                         makecmap_options = list('colFn'=colourmap))

@@

```{r, fig.cap="Snapshot", rgl=TRUE, format="png", eval=TRUE, fig.show='hold'}
#ROI1 
#Figure 1X - Remove white space

parc <- load_parc("Yeo_7")
plot1 <- view_xifti_surface(parc,  
                           legend_embed = T,
                           hemisphere = "right")
```

```{r, fig.cap="Snapshot", rgl=TRUE, format="png", eval=TRUE, fig.show='hold'}
#ROI1 
#Figure 1X - combine with previous figure

@@

parc <- load_parc("Schaefer_400")
ramdom_metric <- rnorm(400)
cii <- move_from_mwall(cii, NA)

parc_vec <- c(as.matrix(parc))
xii_metric <- c(NA, ramdom_metric)[parc_vec + 1]

xii1 <- select_xifti(cii, 1)
xii_metric <- newdata_xifti(xii1, xii_metric)

@@

@@

```{r}
#ROI2

@@

cortex <- ggseg(atlas = glasser,
      .data = base_atlas,
      mapping=aes(fill=Effect), 
      position="stacked",
      colour="black",
      size=.2,
      show.legend = F) + scale_fill_viridis_b()

@@

subcortex <- ggseg(atlas = aseg,
      .data = sub_base_atlas,
      hemisphere = "right",
      mapping=aes(fill=Effect), 
      position="stacked",
      colour="black",
      size=.2,
      show.legend = T) + 
  scale_fill_viridis_b()

cowplot::plot_grid(cortex, subcortex, nrow = 1, rel_widths = c(1,.7))

```
```{r}
#Edge (also add 3d version)
#Fig1X

@@

conmat <- example_weighted_undirected
degree <- rowSums(conmat)[which(rowSums(conmat)!=0)]
brainconn(atlas ="schaefer300_n7", 
          conmat=conmat, 
          node.size = degree/2,
          view="ortho", 
          edge.color.weighted = T,
          background.alpha = 0.4, 
          show.legend = F,
          edge.color = scale_edge_colour_gradient2(low='yellow', mid='orange', high='red'))

>>>>

---
title: "Reproducible and programmatic generation of neuroimaging visualisations"
author: "Sidhant Chopra, Löic Labache, Winnie Orchard, [...]"
date: "09/03/2022"
output:
  pdf_document: default
  word_document: default
  html_document: default
editor_options:
  markdown:
    wrap: 72
---

@@

The `oro.nifti` package allows for the reading, writing, manipulation
and visualisation of voxel-level imaging data of either *nifti*,
*analyze* or *afni* formats. In particular, the `image`, `slice` and
`orthographic` functions allow users to display the desired number of
slices in the desired orientations, while also providing precise control
over overlaid images, colour scales, legend placement and other
aesthetic settings (Fig 1A). Smoothing and re-sampling of images is
sometimes required for visualisation purposes, which can be achieved via
the *`ANTsR`* /`extrantsr` or `fslr` packages.

[view_xifti_volume(xii, interactive = TRUE) in cifti tools]

@@

Statistical parameters can be mapped onto discreet cortical and
sub-cortical regions using the `ggseg` and `ggseg3d` package. These
packages flexibly generates aesthetic renderings of cortical,
sub-cortical and white matter ROIs in 2D and 3D (Fig 1D). Usually,
ROI-level interpretations are dependent on a standardised atlas or
parcellation schemes the brain, accordingly this package and its add-ons
(`ggExtra`) provides a large array of commonly used atlases and some
functionality for users to contribute other atlases. Additionally, being
part of the 'Grammar of Graphics' frame work allows for integration with
packages such as `gganimate`, enabling users to animate and visualise
dynamic changes in spatial and temporal statistics across ROIs. The
`ciftiTools` package also allows for surface and voxel-level
visualisation of ROIs for the cortex and sub-cortex respectively (Fig
1E). Flexibly allowing CIFTI format atlases and ROIs to be displayed.

@@

```{r Figure1 }
library(patchwork)
library(ggplot2)

fig1a <- ggplot() + annotation_custom(grid::rasterGrob(png::readPNG("data/fig1a.png"),
                                                       width=ggplot2::unit(1,"npc"),
                                                       height=ggplot2::unit(1,"npc")))

fig1b <- ggplot() + annotation_custom(grid::rasterGrob(png::readPNG("data/fig1b.png"),
                                                       width=ggplot2::unit(1,"npc"),
                                                       height=ggplot2::unit(1,"npc")))

fig1c <- ggplot() + annotation_custom(grid::rasterGrob(png::readPNG("data/fig1c.png"),
                                                       width=ggplot2::unit(1,"npc"),
                                                       height=ggplot2::unit(1,"npc")))

fig1d <- ggplot() + annotation_custom(grid::rasterGrob(png::readPNG("data/fig1d.png"),
                                                       width=ggplot2::unit(1,"npc"),
                                                       height=ggplot2::unit(1,"npc")))

fig1e <- ggplot() + annotation_custom(grid::rasterGrob(png::readPNG("data/fig1e.png"),
                                                       width=ggplot2::unit(1,"npc"),
                                                       height=ggplot2::unit(1,"npc")))

fig1f <- ggplot() + annotation_custom(grid::rasterGrob(png::readPNG("data/fig1f.png"),
                                                       width=ggplot2::unit(1,"npc"),
                                                       height=ggplot2::unit(1,"npc")))

fig1g1 <- ggplot() + annotation_custom(grid::rasterGrob(png::readPNG("data/fig1g1.png"),
                                                        width=ggplot2::unit(1,"npc"),
                                                        height=ggplot2::unit(1,"npc")))
fig1g2 <- ggplot() + annotation_custom(grid::rasterGrob(png::readPNG("data/fig1g2.png"),
                                                        width=ggplot2::unit(1,"npc"),
                                                        height=ggplot2::unit(1,"npc")))

fig1h <- ggplot() + annotation_custom(grid::rasterGrob(png::readPNG("data/fig1h.png"),
                                                       width=ggplot2::unit(1,"npc"),
                                                       height=ggplot2::unit(1,"npc")))

fig1i <- ggplot() + annotation_custom(grid::rasterGrob(png::readPNG("data/fig1i.png"),
                                                       width=ggplot2::unit(1,"npc"),
                                                       height=ggplot2::unit(1,"npc")))

patchwork <- fig1a/fig1b|(fig1c|fig1f)/fig1d|(fig1e|fig1g2)/fig1g1|fig1h/fig1i

patchwork + plot_annotation(tag_levels = 'A',
                            title = 'Examples of brain visualisations in R')

```{r Fig1a, eval=FALSE}
#Figure 1A
#voxel 1

@@

library(neurobase)
library(ggplotify)

@@

#Plot the effect on the brain template
breaks = seq(.3,.6, by=0.005)

png("data/fig1a.png") 
ortho2(x = template, 
       y = effect, 
       crosshairs = F, 
       bg = "white", 
       NA.x = T, 
       col.y= hotmetal(), 
       xyz = c(70,50,80),
       useRaster = T,
       ycolorbar = TRUE, 
       mfrow = c(1,3)) +
  colorbar(breaks,col = hotmetal(n=length(breaks)-1), 
           text.col = "black", labels = TRUE, maxleft = 0.95) + 
  title(main = "Voxel Statistic Map")
dev.off()

@@

```{r Fig1b, eval=FALSE }
#Figure 1B
#voxel 2

library(neurobase)
library(scales)

@@

png('data/fig1b.png')
overlay(x=robust_window(dti), y=atlas, 
        plot.type = "single", 
        z=c(seq(30,110,10)), 
        col.y =  alpha(rainbow(300), 0.3),
        plane = "axial",  
        useRaster = T, 
        bg = "white",
        NA.x=T, 
        zlim.y =c(1,300), 
        mar = c(0.1, 0.1, 1.5, 0.1)) + #marget of bottom, left, top, right borders
  title("Voxel Atlas Overlay", line = 1.3)
dev.off()

@@

```{r Fig1c , eval=FALSE}
#vertex 1
#Figure 1B
library(rgl)
library(ggplot2)
# Load the package and point to the Connectome Workbench
suppressPackageStartupMessages(library(ciftiTools))

@@

view_xifti_surface(cii, 
                   hemisphere = "left", 
                   view = "both", 
                   idx = 1,
                   colors = viridis::viridis(n = 100),
                   zlim = c(1,1.5),  
                   legend_embed = T,
                   title = "Vertex Statistic Map", 
                   cex.title = 2)

rgl.snapshot("data/fig1c.png")
rgl.close()

@@

```{r Fig1d , eval=FALSE}
#Vertex 2
#Figure 1X - Remove white space

library(fsbrain)

#download_optional_data()
subjects_dir = get_optional_data_filepath("subjects_dir")

@@

vis.subject.morph.native(subjects_dir = subjects_dir, 
                         subject_id = 'subject1', 
                         measure = 'thickness', 
                         hemi='both', 
                         views = 't4',
                         draw_colorbar = "horizontal",
                         cortex_only = T, surface = "white", 
                         makecmap_options = list('colFn'=colourmap))
rgl.snapshot("data/fig1d.png")
rgl.close()

@@

```{r Fig1e , eval=FALSE}
#ROI1 
#Figure 1e 
suppressPackageStartupMessages(library(ciftiTools))
parc <- load_parc("Yeo_7")
view_xifti_surface(parc,  
                   legend_embed = T,
                   hemisphere = "right")

rgl.snapshot("data/fig1e.png")

rgl.close()

```{r Fig1f , eval=FALSE}
#ROI1 
#Figure 1X - combine with previous figure

@@

parc <- load_parc("Schaefer_400")

ramdom_metric <- rnorm(400)

cii <- move_from_mwall(cii, NA)

parc_vec <- c(as.matrix(parc))

xii_metric <- c(NA, ramdom_metric)[parc_vec + 1]

xii1 <- select_xifti(cii, 1)

xii_metric <- newdata_xifti(xii1, xii_metric)

@@

rgl.snapshot("data/fig1f.png")

rgl.close()

@@

```{r Fig1g, eval=FALSE}
#ROI2

@@

cortex <- ggseg(atlas = glasser,
                .data = base_atlas,
                mapping=aes(fill=Effect), 
                position="stacked",
                colour="black",
                size=.2,
                show.legend = F,
                plot.background = "white") + scale_fill_viridis_b()

@@

subcortex <- ggseg(atlas = aseg,
                   .data = sub_base_atlas,
                   hemisphere = "right",
                   mapping=aes(fill=Effect), 
                   position="stacked",
                   colour="black",
                   size=.2,
                   show.legend = T, 
                   plot.background = "white") + 
  scale_fill_viridis_b()

#fig1g <- cowplot::plot_grid(cortex, subcortex, nrow = 1, rel_widths = c(1,.7))

ggsave(filename = 'data/fig1g1.png', plot = cortex, device = 'png', bg = 'white')
ggsave(filename = 'data/fig1g2.png', plot = subcortex, device = 'png', bg = 'white')

```{r Fig1h , eval = F}
#Edge (also add 3d version)
#Fig1X

@@

conmat <- example_weighted_undirected
degree <- rowSums(conmat)[which(rowSums(conmat)!=0)]
fig1h <- brainconn(atlas ="schaefer300_n7", 
                   conmat=conmat, 
                   node.size = degree/2,
                   view="ortho", 
                   edge.color.weighted = T,
                   background.alpha = 0.4, 
                   show.legend = F,
                   edge.color = scale_edge_colour_gradient2(low='yellow', mid='orange', high='red')) 
ggsave(filename = 'data/fig1h.png', plot = fig1h, device = 'png', bg = 'white')

brainconn3D(atlas ="schaefer300_n7", 
            conmat=conmat,
            show.legend = F)
